Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2023-11-29 article.qmd
Introduction

One limitation of these tools is that they are not extensible from Python,
as they are not fully Python packages
but wrap libraries in other languages to do the simulation.
As such, they allow you to "control" the model definition and simulation in a Python script.
But it is not possible to use other packages from the vast Python community to change the way the simulation is carried out.
For instance,
to do a Random Parameter Scan in COPASI,
we must choose between one of 4 predefined distributions,
but cannot use one of the many distributions from `scipy.stats`.

Another weakness is related to the way models are written.
While most tools support standards such as the System Biology Markup Language (SBML) [@huckaSBMLL3V2],
no one is expected to write an SBML file directly.
Tellurium uses a Domain Specific Language (DSL) called Antimony [@smithAntimonyModularModel2009],
which can be translated to SBML.
A DSL allows code to be reused in different programming environments,
but it will not be recognized by Integrated Development Environments (IDEs)
and therefore cannot provide the development experience we have come to expect, such as
syntax highlighting, code completion, refactoring, and static analysis.
This has been recognized by the community,
and an extension for Visual Studio Code has been created [@maTestingDynamicalSystems2021].
However, maintaining it in the future could be burdensome for the smaller systems biology community.
In the case of PySB,
it leveraged the highly dynamic nature of Python to create a DSL within it.
By default, it uses global state to create species and parameters,
without assigning them to Python variables
or adding them explicitly to the model.
But IDEs cannot understand it and, hence, provide help, as this is not the way they expect Python code to be written.

To address these issues,
we developed poincaré and SimBio,
open-source Python packages for defining and simulating systems.
Poincaré allows to define differential equation systems
and SimBio builds on top of it to define reaction networks.
They are focused on providing an ergonomic experience for end-users
by integrating well with IDEs and static analysis tools
through the use of standard modern Python syntax.
They can be extended from Python
since they are coded in pure Python,
which is beneficial for power users.
For instance, it is easy to reuse integrators defined in other packages,
but it is also possible to change the compilation itself
(the )
Their models can be introspected to create other representations,
such as graphs connecting species and/or reactions,
or tables with parameters or equations.
Finally,
they have a modular architecture with a clear separation of concerns,
making it easier to maintain or to contribute new code,
which is beneficial for developers and maintainers.

>>>>

One limitation of these tools is that they are not extensible from Python,
as they are not fully Python packages
but wrap libraries in other languages to do the simulation.
As such, they allow you to "control" the model definition and simulation in a Python script.
But it is not possible to use other packages from the vast Python community to change the way the simulation is carried out.
For instance,
to do a Random Parameter Scan in COPASI,
we must choose between one of 4 predefined distributions,
but cannot use one of the many distributions from `scipy.stats`.
In Tellurium, we cannot choose other Python solvers,
as adding new integrators has to be done in C++.

Another weakness is related to the way models are written.
While most tools support standards such as the System Biology Markup Language (SBML) [@huckaSBMLL3V2],
no one is expected to write an SBML file directly.
Tellurium uses a Domain Specific Language (DSL) called Antimony [@smithAntimonyModularModel2009],
which can be translated to SBML.
A DSL allows code to be reused in different programming environments,
but it will not be recognized by default in Integrated Development Environments (IDEs)
and therefore cannot provide the development experience we have come to expect, such as
syntax highlighting, code completion, refactoring, and static analysis.
This has been recognized by the community,
and an extension for Visual Studio Code has been created [@maTestingDynamicalSystems2021].
However, maintaining it in the future could be burdensome for the smaller systems biology community.
In the case of PySB,
it leveraged the highly dynamic nature of Python to create a DSL within it.
By default, it uses global state to create species and parameters,
without assigning them to Python variables
or adding them explicitly to the model.
But IDEs cannot understand it and, hence, provide help, as this is not the way they expect Python code to be written.

To address these issues,
we developed poincaré and SimBio,
open-source Python packages for defining and simulating systems.
Poincaré allows to define systems of differential equations
and SimBio builds on top of it to define reaction networks.
They are focused on providing an ergonomic experience for end-users
by integrating well with IDEs and static analysis tools
through the use of standard modern Python syntax.
Since they are coded in pure Python,
every part from model definition to simulation can be extended from Python itself.
For example,
easier changes such as reusing integrators defined in other packages,
to more complex ones such as changing the compilation process itself
to leverage some structure in the equations.
Their models can be introspected to create other representations,
such as graphs connecting species and/or reactions,
or tables with parameters or equations.
Finally,
they have a modular architecture with a clear separation of concerns,
making it easier to maintain or to contribute new code,
which is beneficial for developers and maintainers.
